---
title: "PRJNA208535: gflow to gflowui Handoff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PRJNA208535: gflow to gflowui Handoff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette defines a reproducible pipeline to package a PRJNA208535-style analysis as a `gflowui` project.

## Scope

Target workflow:

1. preprocess ASV table (feature reduction),
2. build iKNN graph family for `k = 3:20`,
3. generate 3D layouts via `grip.layout()`,
4. compute median normalized GCV over top-variance ASVs,
5. save graph/layout/optimal-k/CST assets,
6. call `gflowui::register_project(..., overwrite = TRUE)`.

Reference handoff notes can be kept alongside your analysis docs (for example the PRJNA208535 handoff markdown used in VMRC work).

## 1) Preprocess ASV features

```{r, eval = FALSE}
# Input:
#   X_raw: samples x ASVs matrix
#   mt: metadata data.frame with rownames matching samples

# Example reduction: keep top-variance ASVs (tune n_features as needed)
n_features <- 50L
vars <- apply(X_raw, 2, var, na.rm = TRUE)
keep <- order(vars, decreasing = TRUE)[seq_len(min(n_features, ncol(X_raw)))]
X <- X_raw[, keep, drop = FALSE]
```

## 2) Build graph family

```{r, eval = FALSE}
gobj <- gflow::build.iknn.graphs.and.selectk(
  X = X,
  kmin = 3L,
  kmax = 20L,
  method = "edit" # or project-standard method
)

saveRDS(gobj, "results/asv_graphs/iknn.selection.rds")
```

## 3) Create grip layouts for all k

Use the same layout parameters across renderers:

```{r, eval = FALSE}
layout_args <- list(
  dim = 3,
  rounds = 200,
  final_rounds = 200,
  num_init = 10,
  num_nbrs = 30,
  r = 0.1,
  s = 1.0,
  tinit_factor = 6,
  seed = 6
)

dir.create("results/asv_graphs/layouts_3d_rds", recursive = TRUE, showWarnings = FALSE)

for (k in gobj$k.range) {
  adj <- gobj$graphs[[as.character(k)]]$adj.list
  wt <- gobj$graphs[[as.character(k)]]$weight.list
  layout3d <- do.call(grip::grip.layout, c(list(adj_list = adj, weight_list = wt), layout_args))
  saveRDS(layout3d, sprintf("results/asv_graphs/layouts_3d_rds/asv_k%02d_layout3d.rds", k))
}
```

## 4) Compute median normalized GCV criterion

```{r, eval = FALSE}
# Pseudocode:
# - pick top variance ASV outcomes (e.g. first 20)
# - for each k and outcome, run fit.rdgraph.regression + refit.rdgraph.regression
# - compute normalized GCV per outcome
# - aggregate median normalized GCV per k
# - save summary and figure

write.csv(gcv_summary, "results/asv_graphs/optimal_k/median_norm_gcv_by_k.csv", row.names = FALSE)
grDevices::pdf("results/asv_graphs/optimal_k/median_norm_gcv_by_k.pdf", width = 7, height = 5)
plot(gcv_summary$k, gcv_summary$median_norm_gcv, type = "b")
grDevices::dev.off()
```

## 5) Save metadata color assets and optional conditional expectations

```{r, eval = FALSE}
# Example metadata object expected by gflowui color discovery:
mt.asv <- mt[, c("CST", "subCST"), drop = FALSE]
save(mt.asv, file = "data/S_asv.rda")

# Optional condexp assets:
# - fit_files by family/k (type = "fit_files"), or
# - long table with outcome/k/y_fitted/y_observed (type = "long_table")
```

## 6) Register in gflowui

```{r, eval = FALSE}
graph_sets <- list(
  list(
    id = "all",
    label = "ASV Full Graph",
    data_type_id = "all",
    data_type_label = "ASV",
    graph_file = normalizePath("results/asv_graphs/iknn.selection.rds", mustWork = TRUE),
    k_values = 3:20,
    n_samples = nrow(X),
    n_features = ncol(X),
    optimal_k_artifacts = list(
      median_norm_gcv = normalizePath("results/asv_graphs/optimal_k/median_norm_gcv_by_k.pdf", mustWork = TRUE),
      median_norm_gcv_summary = normalizePath("results/asv_graphs/optimal_k/median_norm_gcv_by_k.csv", mustWork = TRUE)
    ),
    layout_assets = list(
      presets = list(renderer = "rglwidget", vertex_layout = "sphere", vertex_size = "1x", color_by = "CST"),
      grip_layout_params = layout_args,
      grip_layouts = lapply(3:20, function(k) list(
        id = sprintf("k%02d", k),
        label = sprintf("k=%d grip.layout", k),
        k = k,
        path = normalizePath(sprintf("results/asv_graphs/layouts_3d_rds/asv_k%02d_layout3d.rds", k), mustWork = TRUE),
        source = "grip.layout"
      ))
    ),
    color_assets = list(
      metadata_file = normalizePath("data/S_asv.rda", mustWork = TRUE),
      metadata_object = "mt.asv",
      vector_columns = c("CST", "subCST"),
      preferred_order = c("CST", "subCST"),
      labels = c(CST = "CST", subCST = "subCST")
    )
  )
)

gflowui::register_project(
  project_root = normalizePath(".", mustWork = TRUE),
  project_id = "prjna208535",
  project_name = "PRJNA208535",
  profile = "custom",
  scan_results = FALSE,
  graph_sets = graph_sets,
  condexp_sets = list(),   # fill when available
  endpoint_runs = list(),  # fill when available
  defaults = list(graph_set_id = "all"),
  overwrite = TRUE
)
```

## Notes

- Keep layout generation deterministic (`seed = 6`) to avoid renderer drift.
- `HTML`, `RGL`, and `Plotly` should all read the same saved `grip.layout()` coordinates.
- Add new color variables through `color_assets$vector_columns`; `gflowui` will expose them in `Color by`.
